---
title: "Data Wrangling Activity"
author: "Your Name"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
library(palmerpenguins)
library(tidyverse)
```


## Goals for the Activity  

1. Use `select()`, `arrange()`, and `filter()` to answer some questions about the penguins.   
2. Connect reorganized data and graphics together.   
3. Learn a few other `dplyr` functions such as `relocate()` and `rename()`.  


Remember to use your resources, such as the slides, readings, and cheatsheets for the `dplyr` package.  

## Penguin Data 

We have used this data in previous activities (and we will continue to do so for a few more activities), but here we can use the `dplyr` function `glimpse()` to remind ourselves what it looks like.  

```{r}
glimpse(penguins)
```



## Practice with `filter()`  

Recall that `filter()` allows us to select specific characteristics via row values within the data.  In addition, we can use the `%>%` (pipe) to connect our data to the function.  

For example, here is the code to filter out just the Chinstrap penguins from the data

```{r}
penguins %>% 
  filter(species == "Chinstrap") 
```

Look at the number of rows in our filtered data frame - how many of the penguins in our data were "Chinstrap"?

> Your Answer


We can then connect our filtered data to a graphic to plot only the remaining data values.  
```{r}
penguins %>% 
  filter(species == "Chinstrap") %>% 
  ggplot(aes(x = bill_length_mm, y = sex, fill = sex)) + 
  #remember to switch to `+` once you start the ggplot2 code
  geom_boxplot() +  #create boxplot
  guides(fill = "none") +  #remove legend generated from fill =
  theme_minimal() + #change theme
  labs(x = "Bill Length (mm)",
       y = "Sex of Penguin",
       title = "Bill Length of Chinstrap Penguins by Sex") + #add labels and title
  scale_fill_viridis_d(begin = 0.25, end = 0.85) #modify default colors
```

### Your Turn with `filter()`  

Find the following information: 

Create a data frame with just Adelie Penguins

```{r}
penguins %>% 
  filter(species=="Adelie")
```


Create a data frame with both Adelie and Chinstrap penguins

```{r}
penguins %>% filter(species=="Adelie" | species=="Chinstrap")
```


Create a data frame with just female Adelie penguins

```{r}
penguins %>% filter(species=="Adelie", sex=="female")
```


Create a data frame of female Adelie penguins with flipper length greater than 190mm.  

```{r}
penguins %>% filter(sex=="female", species=="Adelie", flipper_length_mm>"190")
```


Make a histogram of flipper length for female Adelie penguins with flipper length greater than 190mm. Be sure to include proper labels and titles for the histogram to clearly communicate the data represented.

```{r}
penguins %>% 
  filter(species == "Adelie", sex=="female", flipper_length_mm>"190") %>% 
  ggplot() + 
  geom_histogram(aes(x = flipper_length_mm)) +
  guides(fill = "none") + 
  theme_classic() + 
  labs(x = "Flipper Length (mm)",
       y = "Number of Penguins",
       title = "Flipper length of Female Adelie Penguins") + 
  scale_fill_viridis_d(begin = 0.25, end = 0.85) 
```


## Practice with `select()` 

Recall that `select()` allows us to select specific columns within the data.  In addition, we can use the `%>%` (pipe) to connect our data to the function.  

For example, here is the code to select just the species and sex variables: 

```{r}
penguins %>% 
  select(species, sex)
```

### Your Turn with `select()`

Find the following information: 

Select the columns that contain species, body mass, and sex

```{r}
penguins %>% select(species, body_mass_g, sex)
```


Select the columns that end with "mm"

```{r}
penguins %>% select(ends_with("mm"))
```


Select all columns excluding year 

```{r}
penguins %>% select(-c(year))
```


## New function `relocate()`

We can use `relocate()` to move columns around, without messing with rows or groups. Here are some useful tips:  

- Use `.before` or `.after` to move a column to before or after another (by name or class)  
- If a single column is within the function (e.g. `relocate(col_A)`), that column is moved to the front  

For example, the following will move `year` to the front 
```{r}
penguins %>% 
  relocate(year)
```

Or we could move `year` after `island`
```{r}
penguins %>% 
  relocate(year, .after = island)
```

### Your Turn with `relocate()`

Move sex after species 
```{r}
penguins %>% relocate(sex, .after=species)
```


Move body mass before bill length
```{r}
penguins %>% relocate(body_mass_g, .before = bill_length_mm)
```


### New Function `rename()`
We can use `rename()` to change the name of one or more columns. Generally, it’ll look something like this:

`dataframe %>% rename(new_name = old_name)`


For example, we can rename `island` to `palmer_island`
```{r}
penguins %>% 
  rename(palmer_island = island)
```

### Your Turn with `rename()`

Change the name of year to `study_yr` and the name of body mass to `mass`

```{r}
penguins %>% rename(study_yr=year)
```






## Practice with `arrange()` 
We can use `arrange()` to reorder your data based on the values of specific variable or variables from smallest to largest. The addition of `desc()` can reverse the order from largest to smallest.  

For example, we can arrange the penguins by body mass
```{r}
penguins %>% 
  arrange(body_mass_g)
```

Remember we can also combine multiple functions in a pipe
```{r}
penguins %>% 
  filter(island == "Dream") %>% 
  arrange(body_mass_g) %>% 
  select(species, sex, body_mass_g)
```

### Your Turn with `arrange()`

Use the above functions to answer the following questions:

What is the sex and species of the heaviest penguin in the data?

```{r}
penguins %>%
  arrange(body_mass_g)

#Heaviest is the Gentoo species on the Biscoe island, at about 6300g
```


What is the sex and body mass of the heaviest Chinstrap penguin?

```{r}
penguins %>%
  arrange(body_mass_g) %>% 
  select(sex, species, body_mass_g) %>% 
  filter(species=="Chinstrap")

#Male Chinstrap, 4800g
```


What is the sex and body mass of the smallest Chinstrap penguin on Dream island?

```{r}
penguins %>%
  filter(species=="Chinstrap", island=="Dream") %>% 
  select(sex, body_mass_g) %>% 
  arrange(body_mass_g)

#Female, 2700g
```


## Wrap-Up
The functions of `dplyr` allow you to quickly and efficiently wrangle your data based on various characteristics. Once you have finished with the exercises, knit the file the HTML and submit all files.


